Estimating length of SARS-CoV-2 from doubly interval censored data with missed infections

### 2024-03-12 paper-surv-analysis.tex
initial go at paper on survival analysis

\documentclass[thesis.tex]{subfiles}

\begin{document}
\ifSubfilesClassLoaded{

% \section{Definitions}
% \todo[inline]{These will be defined in the introductory chapters, but are here so that this draft is understandable}
% \begin{itemize}
% \item
%   We assume that, for all episodes $i$, $D_i$ is iid and independent
%   of the time of the infection. Define the survival function
%   $\prob(D_i \geq t \mid B_{j(i)} = b, \theta) = \prob(D_i \geq t \mid \theta) = S_\theta(t)$,
%   where $\theta$ are the parameters controlling the survival
%   distribution (the discussion in this document is valid regardless of
%   the model specified for $S_\theta$ and hence we consider $\theta$
%   as an arbitrary vector of parameters).
% \item
%   The beginning of episode $i$ is known to occur in the interval
%   $[l_{j(i)}^{(b)}, r_{j(i)}^{(b)}]$, and similarly for the end of the infection
%   in $[l_{j(i)}^{(e)}, r_{j(i)}^{(e)}]$.
% \end{itemize}

\setcounter{chapter}{4}
}

\chapter{Estimating duration with infrequent testing} \label{perf-test}

When estimating incidence from prevalence, the mean duration, not its median, is the most important feature of the duration distribution~\autocite{freemanPrevalence}.
Duration distributions tend to be right-skewed.
Therefore, estimates of the mean are sensitive to the tail of the duration distribution.

In \cref{E-ATACCC} I estimated the duration distribution from the ATACCC dataset.
However, this study suffered from a small sample size and only 20 days of follow-up, enabling estimation of the tail.
Therefore, model assumptions drive the size of the estimated tail.

In this chapter, I develop a method to estimate the duration distribution using CIS data.
The CIS sampling is less frequent, but with a large sample size and long follow-up.

I take a Bayesian survival analysis approach to model the coarse sampling of the CIS (\cref{perf-test:sec:problem}).
I use a simulation study (\cref{perf-test:sec:simulation-study}) to evaluate a range of priors for the survival distribution (\cref{perf-test:sec:parameters-priors}).
These include a prior to combine the information in the CIS with the results of \cref{E-ATACCC}.
Reasonable priors successfully recover the true distribution.
However, the method must be able to account for false negatives, discussed in \cref{E-imperf-test}.

\section{Problem description} \label{perf-test:sec:problem}

The CIS tests each individual infrequently, either weekly or monthly (see \cref{E-intro:sec:cis}).
The infrequent testing complicates analysis of the data in two ways.
First, infection episodes can go undetected.
Second, the infection episode start and end times are known imprecisely.
This section explains these issues. The rest of the chapter develops a method to tackle  them.

@@

The first complication in the analysis is that infection episodes can be undetected.
Undetected infection episodes are episodes with no associated positive tests.
Therefore, there is no information directly observed about the episode; even the number that occur is unobserved.
They occur when an episode starts and ends between tests (see \cref{perf-test:fig:truncation}).
Undetected episodes are important because they are shorter than detected episodes.
To see why this is the case, consider the undetected infection in the bottom of \cref{perf-test:fig:truncation}.
The infection episode depicted begins at time 33 and has a 10-day duration.
If the duration had been 24 days or more, the individual would still be detectable at their next test and detected then.
CIS has many undetected infections.
This is because the interval between tests in the same individual (up to 28 days, see \cref{E-intro:sec:cis}) is longer than most infections (median 16 days, see \cref{E-ATACCC:sec:results}).
\begin{figure}
  \makebox[\textwidth][c]{\includegraphics[width=1.2\textwidth]{cis-perfect-testing/truncation}}
  \caption[Undetected episodes in CIS data]{An unknown number of infection episodes in CIS are undetected. Consider the two individuals shown here: we collect identical data for them (a series of negative tests) yet the top individual was never infected and the bottom individual was. See the main text for why this is an issue. \label{perf-test:fig:truncation}}
\end{figure}

\subsection{Unobserved episode beginning and end times} \label{perf-test:sec:interval-censoring}

The second complication is that the exact beginning and end times of detected infection episodes are unknown.
The imprecision in the beginning of the infection episode arises because the only observation that episode $j$ has started is that an individual was negative at one test, on day $l_j^{(b)}-1$ (the minus one is because the first day that the infection could have begun is the day after the negative test), and positive at the next test, on day $r_j^{(b)}$.
Similarly, the end of the infection episode is bounded by the final positive test on day $l_j^{(e)}$ and the following negative test on day $r_j^{(e)}+1$ (the plus one is because the last day that the infection could have ended is the day before the negative test).
\Cref{perf-test:fig:double-interval-censor} shows this issue graphically.
\begin{figure}
  \centering \includegraphics{cis-perfect-testing/double-interval-censor}
  \caption[Double-interval censoring in CIS data]{Episodes data in CIS is double interval censored, meaning that both the start and end of the episode are only known up to an interval. Demonstrated here by a participant who is recorded negative at time 7 to positive at time 14, bounding the start of the episode; similarly, changing from positive at time 28 to negative at time 56 bounds the end of the episode. \label{perf-test:fig:double-interval-censor}}
\end{figure}

\subsection{Approach} \label{perf-test:sec:approach}

I base my analysis on the 4,800 episodes that were first detected between 10th Oct 2020 and 6th Dec 2020 inclusive with negatives bounding the start and end time of the episode.
That is, the individual tested negative before and after the episode.
Without a negative before the episode there is no lower bound on the episode's start time.
It could be arbitrarily long and there is little information on its length.
A negative test follows each episode unless the individual was lost to follow-up before their episode ended.
This is rare.
Episodes removed in this step are incorporated in the adjustment for undetected episodes.

\section{Survival analysis}

In survival analysis, the event at the start of the time of interest is the \emph{initiating event}, and the event at the end of the time of interest is the \emph{terminating event}.
It is used in many areas, including how long a machine lasts before breaking down or how long a patient spends in hospital after a treatment.
Here, I am estimating the distribution of the duration of positivity.
The initiating event is the episode's start; the terminating event is the episode's end.

@@

The time of the initiating and/or terminating events are not known exactly.
This is known as \emph{censored} in the survival analysis literature.
Censored events are present in almost all survival analyses.
An event is \emph{left censored} if it is known to occur before a certain time.
An event is \emph{right censored} if it is known to occur after a certain time.
An event is \emph{interval censored} if it is known to occur within an interval.
%Left and right censoring can be viewed as special cases of interval censoring where the lower bound is negative infinity or the upper bound is positive infinity respectively.
As explained in \cref{perf-test:sec:problem}, the initiating and terminating events are only bounded, and hence interval censored.
The data is \emph{doubly censored} because both the initiating and terminating events are censored.
To handle doubly censored data, the times need to be jointly modelled~\autocite[and references therein]{liSemiparametric}.
In CIS data both the initiating and the terminating events are interval censored.
Therefore, the CIS data is \emph{double interval censored}.
\Textcite{sunAnalysis,bogaertsSurvival} review methods for double interval censored data.

Few studies look at the combination of both double interval censoring and undetected (or truncated) events.
This is especially true within the human biostatistical literature.
The literature does include theoretical frameworks which include the double censored and truncated case~\autocite{turnbullEmpirical,dempsterMaximum}, yet these frameworks have only been applied when the terminating event was either uncensored or right censored~\autocite{sunEmpirical,bacchettiNonparametric}.
These studies \autocite[and elsewhere, e.g.:][]{shenNonparametric} take a conditional likelihood approach.
They condition on the truncation (a selection effect) and the period where the initiating event occurs.
They condition on the initiating event time because it provides negligible information on the distribution of interest~\citePersonalComms{Nick Jewell}.
This is not the case for the CIS.
In CIS, infection episodes are more likely to be detected when an individual is in the weekly testing phase.
Therefore, the probability of an infection episode being detected conditional on being infected is higher than the overall probability of being detected.

Censored and truncated data are normal for estimation of the survival time of bird and insect nests~\autocite{heiseyABCs}.
\textcite{heiseyModelling} developed a framework for handling this situation.
The framework is general, allowing for arbitrary censoring patterns and double interval censoring.
They apply it to a situation the visit times are global (all nests could be detected on the same days), simplifying the likelihood.
I adapt this framework for CIS, when each individual has their own visit schedule (see section \cref{perf-test:sec:model}).

I use Bayesian inference for two reasons.
First, frequentist estimators are poorly developed and their properties (\eg consistency and convergence rate) are poorly understood~\autocite{sunAnalysis,dengNonparametric}.
Second, a prior can include information from previous studies, such as from \cref{E-ATACCC}.
Alternatively, the prior can favour a smoothly changing hazard.
I explore priors in section 1.4.
These prior structures overcome issues caused by the undetected episodes ~\autocite{caoBias}.

Bayesian methods have been used for nest survival before~\autocite{heBayesiana,heBayesian,caoModeling} by augmenting the data with the unobserved times of infection.
The augmentation creates a parameter per detected infection.
This increases the dimensionality of the problem greatly, making it computationally infeasible within the SRS.
I develop a novel Bayesian framework without explicit data augmentation.
This method scales to the large number of episodes present in CIS.

\section{Modelling the duration}\label{perf-test:sec:model}

In this section, I develop a model accounting for double interval censoring and undetected episodes.
This approach extends the framework of \textcite{heiseyModelling} to include individuals without detected episodes.
This framework is itself based long-standing methodology~\autocite{dempsterMaximum,turnbullEmpirical}.
I augment the data set with the full number of episodes, including those that are undetected.

@@

% $\ntot$ is generally a nuisance parameter, however, its posterior distribution can be useful for diagnostic purposes (as in X\todo{ref where I use this in the next chapter}).
% Its posterior can be reconstructed using a posterior sample of $\theta$.
% For each posterior sample of $\theta$, sample from the full conditional of $\ntot$ to give the joint posterior of $\ntot$ and $\theta$.
% The full conditional of $\ntot$ is given by:
% \begin{align}
% p(n_\text{tot} \mid \ncis, \theta)
% &\propto p(\ntot \mid \theta) p(\ncis \mid \theta, \ntot) \\
% &\propto p(\ntot \mid \theta) \frac{\ntot!}{(\ntot - \ndet)!} \pnodet^{\ntot - \ndet} &\text{by \cref{perf-test:eq:augmented-likelihood}} \\
% &\propto \frac{\Gamma(r + \ntot)}{\ntot!} \left( \frac{\mu}{r + \mu} \right)^{\ntot} \frac{\ntot!}{(\ntot-\ndet)!} \pnodet^{\ntot} \\
% &= \frac{\Gamma(r + \ntot)}{(\ntot-\ndet)!} \left( \frac{\mu \pnodet}{r + \mu} \right)^{\ntot}  \\
% &= \frac{\Gamma((r + \ndet) + (\ntot- \ndet))}{(\ntot-\ndet)!} \left( \frac{\mu \pnodet}{r + \mu} \right)^{\ntot-\ndet}.
% \end{align}
% Comparing this expression to the pmf of a negative binomial, we find that $\nnodet = \ntot - \ndet$ is distributed negative binomial with size parameter $r+\ndet$ and probability parameter $(r + \mu(1 - \pnodet))/(r + \mu)$.
%The mean of this distribution is $(r+\ndet)\mu \pnodet/(r+\mu(1-\pnodet))$.

@@

\section{Parameterisation and priors for the survival function} \label{perf-test:sec:parameters-priors}

In this section I specify the form of $S_\theta(t)$.
% As frequently observed in the literature~\autocite[e.g.][]{heBayesian}, it is more convenient to operate with the hazard than the survival or probability mass function directly.
% The mass function must sum to one.
% The survival must be monotonically decreasing
% But, there are no constraints across multiple hazards.
% This would complicate the inference process in a way that cannot be removed through a simple transformation.
% The only constraint on the hazard is to be in the interval $[0, 1]$, since they are probabilities.
% The [0, 1] interval is easily be mapped to the unconstrained reals using a logit transformation (where $\logit(x) = \log(x/(1-x))$).
Priors cannot be vague on both the hazard and the survival, the implications of which I consider in what follows.
Informative priors are also attractive to allow the incorporation of the estimates from \cref{E-ATACCC}.

\subsection{Independent priors} \label{perf-test:sec:independent-priors}
Take the standard assumption of independent priors on each parameter.
Each hazard, $\lambda_t$, is a (conditional) probability, and hence a Beta distribution is a natural choice.

Uninformative Beta priors on the hazard are problematic in this context.
An uninformative prior would be of the form $\lambda_t \dist \BetaDist(\alpha, \alpha)$.
Commonly, $\alpha$ is chosen to be $0.5$ (Jeffreys' Prior) or $1$ (a uniform prior).
Even though these priors are uninformative on the hazard, they become highly informative on the survival. 
Specifically, they tend to favour shorter survival times (see \cref{perf-test:fig:flat-prior}).
An intuitive explanation of this can be drawn from the fact that priors of this form have an expected value of 0.5.
The expected survival time, $\E \left( \prod_{i=1}^{t-1} (1-\lambda_t) \right)$, is therefore equal to $0.5^{t-1}$, a quantity that declines rapidly.
This prior expresses extreme scepticism in estimates from previous studies, such as \cref{E-ATACCC} and the meta-analysis of \textcite{cevikShedding}.
\begin{figure}
  \centering \includegraphics{cis-perfect-testing/flat-prior}
  \caption[Uninformative priors for the hazard]{The density of Beta(1, 1) flat prior and Beta(0.5, 0.5) uninformative priors (left) with the prior predictive survival time (right). The implied prior on the survival time is very short when using either prior on the hazard at each time. \label{perf-test:fig:flat-prior}}
\end{figure}

Instead, I propose a weakly informative (or vague) prior Beta(0.1, 1.9), which has mean 0.05 and minimal information.
The amount of information in a Beta distribution is related to the sum of its parameters.
Here, the sum equals 2, the same as the flat prior case.
The central 95\% probability mass of Beta(0.1, 1.9) is 0.00--0.47.
The central estimate, of 0.05, is in line with previous estimates that the median duration is in the range 15--20 days~\autocite{cevikShedding}.
This prior gives a very vague prior predictive distribution on the survival time (see \cref{perf-test:fig:vague-prior}).
\begin{figure}
  \centering \includegraphics{cis-perfect-testing/vague-prior}
  \caption[Vague prior for the hazard]{The density of Beta(0.1, 1.9) vague prior (left) and the prior predictive survival time (right). The prior predictive survival time is very vague using a Beta(0.1, 1.9) prior. Note the different x-axis to \cref{perf-test:fig:flat-prior}. \label{perf-test:fig:vague-prior}}
\end{figure}

\subsection{Smoothing priors}

Smoothing priors encode the biological consideration that the hazard should not vary significantly from day-to-day.
It is the Bayesian equivalent of a penalised likelihood, used frequently in the non-parametric frequentist setting~\autocite[e.g.][]{bacchettiNonparametric}.
Many possible forms of smoothing priors exist, including splines, Gaussian Processes, and random walks.
I use a second-order random walk (RW2) because it is simple and produces sensible prior predictive distributions.

>>>>

\documentclass[12pt, letterpaper]{article} % Sets the font size to 12pt and paper size to letter

\usepackage[margin=1in]{geometry} % Sets the margins to 1 inch all around
\usepackage{setspace} % Allows for double spacing and setting line spacing
\setlength{\marginparwidth}{2cm}

\usepackage{graphicx}
\usepackage[utf8]{inputenc}

\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{floatpag}
\usepackage{bm}
\usepackage{algorithm2e}
\usepackage{csquotes}
\usepackage[T1]{fontenc}
\usepackage{hyperref}
\usepackage{textcomp} % provide symbols
\usepackage{todonotes}
\usepackage{xr}
\usepackage{pdflscape}
\usepackage{afterpage}
\usepackage{caption}
\usepackage{numprint}
% \usepackage{microtype}
\usepackage[USenglish]{babel}
\usepackage[nameinlink]{cleveref}
\usepackage[backend=biber,style=apa,natbib=true]{biblatex} % Load biblatex with the desired settings
\addbibresource{references.bib}
\addbibresource{refs-custom.bib}
% no URLs
\AtEveryBibitem{
  \clearfield{url}
  \clearfield{doi}
}

\captionsetup{font=footnotesize} % Adjust the caption font size as necessary for readability

% \usepackage{lineno} % Include the package for line numbering
% \linenumbers % Activates line numbering for the document

% Diagrams
\usepackage{tikz}
\usetikzlibrary{arrows,positioning,automata,fit}

% Macros for common abbreviations to get the spacing right
% See: https://stackoverflow.com/questions/3282319/correct-way-to-define-macros-etc-ie-in-latex
\usepackage{xspace}
\makeatletter
\DeclareRobustCommand\onedot{\futurelet\@let@token\@onedot}
\def\@onedot{\ifx\@let@token.\else.\null\fi\xspace}
\def\eg{e.g\onedot} \def\Eg{\emph{E.g}\onedot}
\def\ie{i.e\onedot} \def\Ie{\emph{I.e}\onedot}
\def\cf{c.f\onedot} \def\Cf{\emph{C.f}\onedot}
\def\etc{etc\onedot} \def\vs{\emph{vs}\onedot}
\def\wrt{w.r.t\onedot} \def\dof{d.o.f\onedot}
\def\etal{et al\onedot}
\makeatother

% Generic maths commands
\def\reals{\mathbb{R}}
\def\nats{\mathbb{N}}
\def\sampSpace{\mathcal{X}}
\def\dist{\sim}
\DeclareMathOperator{\E}{\mathbb{E}}
\DeclareMathOperator{\V}{\mathbb{V}}
\DeclareMathOperator{\I}{\mathbb{I}}
\DeclareMathOperator{\prob}{\mathbb{P}}
\DeclareMathOperator{\p}{\pi}
\DeclareMathOperator{\var}{\mathbb{V}}
\DeclareMathOperator{\indicator}{\mathbb{I}}
\DeclareMathOperator{\cov}{Cov}
\DeclareMathOperator{\cor}{Cor}
\DeclareMathOperator{\logit}{logit}
\DeclareMathOperator{\Ber}{Bernoulli}
\DeclareMathOperator{\Bin}{Binomial}
\DeclareMathOperator{\Poi}{Poisson}
\DeclareMathOperator{\BetaDist}{Beta}
\DeclareMathOperator{\Exponential}{Exponential}
\DeclareMathOperator{\NBr}{NegBin}
\newcommand{\NBc}{\NBr_{c}}
\newcommand{\NBs}{\NBr_{s}}
\DeclareMathOperator{\BB}{BetaBin}
\DeclareMathOperator{\GamDist}{Gamma}
\DeclareMathOperator{\MN}{Multinomial}
\DeclareMathOperator{\N}{N}
\DeclareMathOperator{\LN}{LN}
\DeclareMathOperator{\LKJ}{LKJ}
\DeclareMathOperator{\expit}{expit}
\newcommand\matr{\bm}
\newcommand\set{\mathcal}
\renewcommand{\vec}{\bm}
\newcommand{\ssep}{:}

% Thesis-specific maths commands
\newcommand{\dmax}{d_\text{max}}
\newcommand{\psens}{p_\text{sens}}
\newcommand{\ntot}{n_\text{tot}}
\newcommand{\ndet}{n_\text{d}}
\newcommand{\nnodet}{n_\text{u}}
\newcommand{\pnodet}{p_\text{u}}
\newcommand{\Npop}{N_\text{pop}}
\newcommand{\Ncis}{N_\text{CIS}}
\newcommand{\ncis}{\vec{n_\text{CIS}}}
\newcommand{\na}{\vec{n_\text{a}}}
\newcommand{\pcis}{\vec{p_\text{CIS}}}
\newcommand{\sched}{\mathbb{T}}
\newcommand{\nsched}{n_{\sched}}
\newcommand{\rate}{\mathfrak{R}}
\newcommand{\R}{\mathcal{R}_{0}}
\renewcommand{\Re}{\mathcal{R}_{e}(t)}
\newcommand{\Rc}{\mathcal{R}_{c}(t)}
\newcommand{\vlod}{v_\text{lod}}
\newcommand{\inform}{{_{\text{inform}}}}

\begin{document}

\title{Estimating length of SARS-CoV-2 from doubly interval censored data with missed infections}
\author{Blake, TBC, Birrell, De Angelis}
% \date{\today}
\maketitle

\begin{abstract}
Abstract goes here.
\end{abstract}

The format requirements from JASA are at \url{https://www.tandfonline.com/journals/uasa20}.
The most important parts are reproduced here.
\begin{itemize}
    \item Should be written with the following elements in the following order: title page; author footnote; abstract; keywords; article text (table(s); figures); acknowledgments; appendices; references
    \item Should be no more than 35 pages, including the abstract, figures, tables, and references. Appendices, proofs, and other supporting material should be placed in a separate supplement file (anonymized for review)
    \item Should contain an unstructured abstract of 200 words.
    \item Should contain between 3 and 5 keywords. Read making your article more discoverable, including information on choosing a title and search engine optimization.
    \item JASA requires that all papers be formatted for 8 1/2 x 11-inch paper, one side only.
    \item 12 point, double-spaced font (which we define as 26 lines of text per page)
    \item the average JASA paper is about 30 manuscript pages (with the above page format and including all appendices, references, tables, and figures), and it is very uncommon for papers much longer than about 35 pages to appear in JASA. 
\end{itemize}

Notes / thoughts on current version
\begin{itemize}
    \item Some of the derivations can be moved to a supplement (cut 2--3 pages).
    \item Need to rearranage the ATACCC stuff --- add some details of what I did (thesis ch 4) and move stuff from prior info section into supplement.
    \item Probably don't need some of the figures (maybe 5--8)?
    \item Should the sensitivity analyses (figs 13/14 be in the main text)?
    \item The above changes make this about the right length.
\end{itemize}

Table of contents, cut for final version but useful for now to compare lengths of sections.

\tableofcontents

\newpage
\setstretch{1.7} % Fully double-spaced font. Adjust the value if needed to achieve 26 lines per page

\section{Introduction}

\todo[inline]{Rough outline of this part, which motivates the problem. }

The distribution of duration an individual tests RT-PCR positive with SARS-CoV-2 is important for blah blah reasons.

Investigations into this distribution have been predominantly conducted in highly selected population, \eg healthcare works.
Cite and expand.

The ideal study design is frequent, long-term, repeated testing of a large, representative sample of the population.
No study exists with this design.

Here, we use external information from ATACCC to inform estimates with CIS.
We estimate the distribution of duration an individual tests RT-PCR positive with SARS-CoV-2.

The analysis of CIS is challenging because it is doubly interval censored, misses infections, and contains false negative results.
We develop a Bayesian survival analysis model to estimate the distribution of duration, accounting for all these issues.

\section{Study design and data}

The CIS~\autocite{CIS} was setup in April 2020.
The dataset is globally unique in providing a representative, longitudinal, and large-scale study across the pandemic.
% Initially, it was limited to England, but expanded to cover the whole of the UK in September 2020.
Enrolment was continuous until 31st Jan 2022, with data collected until 13th Mar 2023~\autocite{weiRisk}.

The CIS had a household-based design inviting all individuals aged 2 and over from selected households.
Once invited, an enrolment swab would be taken at the first visit followed by 4 further weekly visits (giving a total of 5 swabs on days 0, 7, 14, 21, 28 relative to enrolment) after which visits were monthly.
As well as a swab to test for SARS-CoV-2, participants were asked to complete a questionnaire, and some participants were asked to provide blood samples.
In reality, visits were often not on this precise schedule, and occasionally missed.
A full description of the study can be found in the study protocol~\autocite{cisProtocol}.

Positive tests are grouped into \emph{infection episodes} which are assumed to originate from the same infection (procedure described in \cref{sec:episodes}).
I base my analysis on the 4,800 episodes that were first detected between 10th Oct 2020 and 6th Dec 2020 inclusive with negatives bounding the start and end time of the episode.
That is, the individual tested negative before and after the episode.
Without a negative before the episode there is no lower bound on the episode's start time.
It could be arbitrarily long and there is little information on its length.
A negative test follows each episode unless the individual was lost to follow-up before their episode ended.
This is rare.
Episodes removed in this step are incorporated in the adjustment for undetected episodes.

For privacy reasons, the CIS data is stored in the ONS's SRS (Secure Research Service).
The SRS is a Trusted Research Environment (TRE) that allows researchers to access sensitive data in a secure environment.
The SRS must be used to perform analyses using individual-level data.
Therefore, computationally efficient approaches must be taken to this work.

\section{Problem description}

The infrequent testing used by the CIS complicates analysis of the data in two ways.
First, infection episodes can go undetected.
Second, the infection episode start and end times are known imprecisely.
This section explains these issues.

@@

The first complication in the analysis is that infection episodes can be undetected.
Undetected infection episodes are episodes with no associated positive tests.
Therefore, there is no information directly observed about the episode; even the number that occur is unobserved.

@@

They occur when an episode starts and ends between tests (see \cref{perf-test:fig:truncation}).
Undetected episodes are important because they are shorter than detected episodes.
To see why this is the case, consider the undetected infection in the bottom of \cref{perf-test:fig:truncation}.
The infection episode depicted begins at time 33 and has a 10-day duration.
If the duration had been 24 days or more, the individual would still be detectable at their next test and detected then.
CIS has many undetected infections.
This is because the interval between tests in the same individual (up to 28 days, see \cref{E-intro:sec:cis}) is longer than most infections (median 16 days, see \cref{E-ATACCC:sec:results}).
\begin{figure}
  \makebox[\textwidth][c]{\includegraphics[width=1.2\textwidth]{cis-perfect-testing/truncation}}
  \caption[Undetected episodes in CIS data]{An unknown number of infection episodes in CIS are undetected. Consider the two individuals shown here: we collect identical data for them (a series of negative tests) yet the top individual was never infected and the bottom individual was. See the main text for why this is an issue. \label{perf-test:fig:truncation}}
\end{figure}

\subsection{Double interval censoring} \label{perf-test:sec:interval-censoring}

The second complication is that the infections are double interval censored.
The imprecision in the beginning of the infection episode arises because the only observation that episode $j$ has started is that an individual was negative at one test, on day $l_j^{(b)}-1$ (the minus one is because the first day that the infection could have begun is the day after the negative test), and positive at the next test, on day $r_j^{(b)}$.
Similarly, the end of the infection episode is bounded by the final positive test on day $l_j^{(e)}$ and the following negative test on day $r_j^{(e)}+1$ (the plus one is because the last day that the infection could have ended is the day before the negative test).
\Cref{perf-test:fig:double-interval-censor} shows this issue graphically.
\begin{figure}
  \centering \includegraphics{cis-perfect-testing/double-interval-censor}
  \caption[Double-interval censoring in CIS data]{Episodes data in CIS is double interval censored, meaning that both the start and end of the episode are only known up to an interval. Demonstrated here by a participant who is recorded negative at time 7 to positive at time 14, bounding the start of the episode; similarly, changing from positive at time 28 to negative at time 56 bounds the end of the episode. \label{perf-test:fig:double-interval-censor}}
\end{figure}

\subsection{False negatives}

Write something about false negatives being common.

\section{Related work}

Few studies look at the combination of both double interval censoring and undetected (or truncated) events.
This is especially true within the human biostatistical literature.
The literature does include theoretical frameworks which include the double censored and truncated case~\autocite{turnbullEmpirical,dempsterMaximum}, yet these frameworks have only been applied when the terminating event was either uncensored or right censored~\autocite{sunEmpirical,bacchettiNonparametric}.
These studies \autocite[and elsewhere, e.g.:][]{shenNonparametric} take a conditional likelihood approach.
They condition on the truncation (a selection effect) and the period where the initiating event occurs, which provides negligible information on the distribution of interest.
This is not the case for the CIS.
In CIS, infection episodes are more likely to be detected when an individual is in the weekly testing phase.
Therefore, the probability of an infection episode being detected conditional on being infected is higher than the overall probability of being detected.

Censored and truncated data are normal for estimation of the survival time of bird and insect nests~\autocite{heiseyABCs}.
\textcite{heiseyModelling} developed a framework for handling this situation.
The framework is general, allowing for arbitrary censoring patterns and double interval censoring.
They apply it to a situation the visit times are global (all nests could be detected on the same days), simplifying the likelihood.
I adapt this framework for CIS, when each individual has their own visit schedule (see section \cref{perf-test:sec:model}).
This framework is itself based long-standing methodology~\autocite{dempsterMaximum,turnbullEmpirical}.

The inclusion of false negatives into survival analysis is an area of much interest, especially in the application of tests for HIV in infants~\autocite[e.g.][]{brownBayesian,balasubramanianEstimation};
\textcite{piresIntervalMisclassify} provide a comprehensive review of approaches.
However, including false negatives with either doubly censored or truncated data has not been addressed.
The situation here requires both.

\section{Methodology}

\subsection{Survival analysis framework}

First, we consider the situtation without false negatives.
This approach extends the framework of \textcite{heiseyModelling} to include individuals without detected episodes.
I augment the data set with the full number of episodes, including those that are undetected.

@@

@@

\subsection{False negatives}

\subsubsection{Generative model for false negatives} \label{imperf-test:sec:simulate}

\todo[inline]{Probably the wrong place for this subsection, and can be edited for brevity a lot}

In this section, I introduce a probabilistic model for the generation of false negatives.
The simplest model for false negatives is a constant, non-zero probability of returning a negative test when detectable.
That is the test sensitivity, $\psens$, is less than 1.

Under this model, each test result in a detectable individual is an iid Bernoulli random variable with a probability $\psens$ of producing a positive.
As previously, tests in individuals without ongoing infection episodes are always negative.
I modified the simulation study from \cref{E-perf-test:sec:simulation-study} to generate test results in this way.

Following the modification, more single positive episodes are seen at lower values of $\psens$, with a value between 60\% and 80\% reproducing the number of single positive episodes observed in the CIS data (see \cref{imperf-test:fig:sim-single-pos}).
This is expected: a lower $\psens$ means that multiple positive tests are less likely, and hence the probability of a single positive episode is higher.

External information (see \cref{sec:ATACCC-analysis}) gives that $\psens$ is 95\% (95\% CrI: 93--96\%).
The requirement for a much lower value of $\psens$ in the simulation study here suggests this model of false negatives is inadequate.

A better model would be to allow the rate of false negatives to vary over the course of an episode.
False negatives are more likely to occur when the viral load of an individual is low, because there is less virus in their body to sample.
In \cref{E-ATACCC}, this mechanism is incorporated by viewing negative tests as a left-truncation of the observation noise (see \cref{E-ATACCC:sec:observation-modification}).
This observation suggests that a model with a declining test sensitivity as a function of time since infection might be more suitable.

The CIS data provides evidence for a test sensitivity declining over the course of an episode.
\Cref{imperf-test:fig:bounding-cis-sensitivity} demonstrates this by using the following classification for test results.
\begin{enumerate}
    \item All positive results are true positives.
    \item All intermittent negatives are false negatives.
    \item The negative at the end of an episode is a possible false negative.
    \item All other negatives are true negatives.
\end{enumerate}
Here, I am assuming that the probability of two false negative tests at the end of an episode is negligible.
As test sensitivity is the proportion of individuals who should test positive that do so, it can be empirically calculated as number of true positives / (number of true positives + number of false negatives).
Therefore, I calculate the test sensitivity for each day, considering the first positive result in an episode as day 0.
As the latent start day of the infection episode is before the first positive result, day 0 reflects some unknown number of days into the infection episode.
I bound the test sensitivity on each day by assuming either all or none of the possible false negatives (group 3) are in fact false negatives.
% In \cref{imperf-test:fig:bounding-cis-sensitivity} we consider the values produced by assuming the negative following the last positive in an episode could be either a true or false negative as a function of time since the individual was detected (\ie: the first positive test in the episode).
This generates broad bounds, although suggest a declining test sensitivity over the course of an episode (see \cref{imperf-test:fig:bounding-cis-sensitivity}).
Small numbers of tests mean that the series is noisy near the start and the end, such as both bounds being 0 on some days after day 40.
However, the general trend is clear.
\begin{figure}
  \includegraphics{cis-imperfect-testing/test-sens-bound}
  \caption[Bounding test sensitivity using CIS data]{
    Bounding the test sensitivity using CIS data as a function of time since the infection was detected.
  }
  \label{imperf-test:fig:bounding-cis-sensitivity}
\end{figure}

To assess the sensitivity of the procedure to a time-varying test sensitivity, it will be useful to simulate from a model with a varying test sensitivity as a function of days since the start of the infection episode.
Through consideration of \cref{imperf-test:fig:bounding-cis-sensitivity}, and the interaction of false negatives due to viral load in \cref{E-ATACCC}'s results, I propose the following model.
\begin{equation}
  p_\text{sens}(t) = \begin{cases}
    0.9 - \frac{0.9-0.5}{50}t &t \leq 50 \\
    0.5 &t > 50
  \end{cases}
  \label{imperf-test:eq:variable-test-sensitivity}
\end{equation}
where $t$ is the number of days since infection.
Therefore, this model consists of a linear decline in test sensitivity, before a constant minimum point.

I simulate data under this model by, for each test in a detectable individual, calculating $\psens(t-b)$, where $t$ is the day of the test and $b$ is the day the episode began.
The result of the test is then a Bernoulli random variable, independent of all other considerations.
Adapting the simulation study to incorporate this test sensitivity is much more similar to the data (see purple histogram in \cref{imperf-test:fig:sim-single-pos}).
This model does not require low test sensitivities early in the infection, maintaining compatibility with the results in \cref{E-ATACCC}.

\subsubsection{Incorporation into the survival analysis framework}

Now we modify the survival framework to incorporate false negatives.
We use the model with a constant test sensitivity to ensure the likelihood remains tractable.

First, in \cref{imperf-test:sec:modifying-p_ia}, I modify $p_{ia}$, the probability of an episode with test schedule $i$ being admissible, to allow for the episode possibly being longer than observed.
Then, in \cref{imperf-test:sec:modifying-p_iu}, I modify $p_{iu}$, the probability of an episode with test schedule $i$ being undetected, to allow for additional episodes being missed.

\subsection{Modifying \texorpdfstring{$p_{ia}$}{pia}} \label{imperf-test:sec:modifying-p_ia}

I will modify $p_{ia}$ to allow the negative test following the last positive to be a false negative.
If it is a false negative, then I will consider the episode's end right-censored.
However, if it is a true negative, then the episode's end is interval-censored, as previously.
A mixture of these scenarios forms the episode's likelihood contribution, with the mixture probability determined by the test sensitivity.

Similar likelihoods have previously appeared in the literature~\autocite[e.g.][eq.\ (2)]{piresIntervalMisclassify}.
However, this prior work was for singly interval censored data; incorporating the double interval censored nature of the CIS data will involve summing over the possible episode start times.

I start by finding the tests that need consideration.
For tractability and simplicity, I consider only tests between the negative tests providing an upper bound on the length of the episode.
By definition (see \cref{E-biology-data:sec:cis-episodes}), these are the tests between $l_{j(i)}^{(b)} - 1$ and $r_{j(i)}^{(e)} + 1$ inclusive.
They are a subset of $\sched_i$, the set of test times for the individual in which episode $i$ occurs.

For tractability, assume that the negative test bounding the start of the episode, on day $l_{j(i)}^{(b)}-1$, is a true negative.
This assumption is reasonable because, since a positive test follows at $r_{j(i)}^{(b)}$, the negative at $l_{j(i)}^{(b)}-1$ is likely early in the infection and the test sensitivity is high early in an infection.
Therefore, this is unlikely to be a false negative because, if the individual was detectable at this time, they are in a period with high test sensitivity.
This true negative, occurs with probability 1, and hence does not contribute to the likelihood.
Denote the remaining tests considered as $\sched'_i = \{ t \in \sched_i \ssep r_{j(i)}^{(b)} \leq t \leq r_{j(i)}^{(e)} + 1 \}$, and their results by $\vec{y_{j(i)}} = \{ y_{j(i)}(t) \ssep t \in \sched'_i \}$ where $y_{j(i)}(t) = 1$ if the test on day $t$ is positive and 0 otherwise.

As I assume that there are no false positives, the infection episode must span at least the period $[r^{(b)}_{j(i)}, l^{(e)}_{j(i)}]$ because it starts and ends with a positive test.
This includes all $t \in \sched'_i$ except $t = r_{j(i)}^{(e)}+1$.
Therefore, the test results at times in $\sched_i'$, except the test at $r_{j(i)}^{(e)}+1$, are either true positives or false negatives.
